Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-06-17 parallel-tractability-J-v9/evaluation.tex
revised the constant time argumentation

We implement a prototype system ParallelDHL for DHL$(\circ)$ materialization
based on Algorithm~$\mathsf{A}_{\text{prc}}$. Since ParallelDHL is
evaluated on a
platform which has limited memory space and a fixed number of processors,
the parallel assumption given in Section~\ref{sec:alg-bsc} does not apply to ParallelDHL.
In the implementation of ParallelDHL, we use a hash function to map any rule instantiation
to the identifier of some processor. Thus, when performing ParallelDHL,
each processor handles a group of rule instantiations.

>>>>

We implement a prototype system ParallelDHL for DHL$(\circ)$ materialization
based on Algorithm~$\mathsf{A}_{\text{prc}}$. Since ParallelDHL is
evaluated on a
platform which has limited memory space and a fixed number of processors,
the parallel condition given in Section~\ref{sec:alg-bsc} does not apply to ParallelDHL.
In the implementation of ParallelDHL, we use a hash function to map any rule instantiation
to the identifier of some processor. Thus, when performing ParallelDHL,
each processor handles a group of rule instantiations.

### 2018-06-17 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
revised the constant time argumentation

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{\text{bsc}}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. This model allows us to introduce \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instantiation in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instantiations.
Under this assumption, a processor can check the applicability of its corresponding
rule instantiation and access the state of an atom occurring in this rule instantiation in constant time.
Algorithm~$\mathsf{A}_{\text{bsc}}$ is then given as follows.\\

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any predicate in $\textbf{I}$, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instantiations in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instantiations is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{\text{bsc}}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$ turns out to be $c_1+\ell\cdot c_2$
where $\ell$ denotes the number of iterations of \ref{alg1:updateG}.

@@

We now analyse the parallel complexity of Algorithm~$\mathsf{A}_{\text{opt}}$.
We use the same symbols $e$, $v$ and $r$ that are used for analyzing the complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$
(see Section~\ref{sec:alg-bsc}). Similarly to the analysis for Algorithm~$\mathsf{A}_{\text{bsc}}$, it can be checked that
the number of processors is polynomial in $|\textbf{I}|$, i.e., $r(|\textbf{I}|e)^v$.
We now consider the computing time of Algorithm~$\mathsf{A}_{\text{opt}}$. It
is obvious that \ref{alg3:addFacts} and
\ref{alg3:halt} occupy constant time (denoted by $c_1$). In \ref{alg3:updateG}, the phases of
computing $S_{\!\textit{\tiny rch}}$ and updating $\mathcal{G}$ also take constant time (denoted by $c_2$)
under the parallel assumption. Recall that the number of constants is at most $|\textbf{I}|e$ (see Section~\ref{sec:alg-bsc}).
Let $w$ and $p$ represent the maximum arity of any predicate and the number of predicates, respectively.
We have that the number of all possible substitutions of atoms is at most $p(|\textbf{I}|e)^w$.
The relation $S_{\!\textit{\tiny rch}}$ consists of pairs of the atoms. Thus, the size of $S_{\!\textit{\tiny rch}}$
is bounded by $p^2(|\textbf{I}|e)^{2w}$. It can be checked that the computing time of the \texttt{NC} algorithm for computing the transitive
closure of $S_{\!\textit{\tiny rch}}$ is poly-logarithmic in the size of $\textbf{I}$, more precisely,
it is upper bounded by $\log^2(p^2(|\textbf{I}|e)^{2w})$ ($=4\log^2(p)+4w^2\log^2(|\textbf{I}|e)+8w\log(p)\log(|\textbf{I}|e)$).
We now have that the total computing time of
Algorithm~$\mathsf{A}_{\text{opt}}$ is $c_1+\ell c_3 + \ell
c_4\log(|\textbf{I}|e) +\ell c_5\log^2(|\textbf{I}|e)$
where $c_3=c_2+4\log^2(p)$, $c_4=8w\log(p)$, $c_5=4w^2$ and $\ell$ denotes the number of iterations of \ref{alg3:updateG}.

>>>>

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{\text{bsc}}$) that constructs a materialization graph for a given datalog program.
The basic computational model is PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. For any datalog program $P=\langle R, \textbf{I}\rangle$, the PRAM model
allows each rule instantiation in $P^*$ corresponding to a unique processor, and each processor accessing the
location of any atom in constant time.\footnote{This is feasible based on the PRAM model and
the parallel prefix computation \cite{LaFi80a}. We refer the readers to the details in \cite{Raymond95}.} 
We use \emph{the parallel condition} to denote the above condition.
On the parallel condition, a processor can check the applicability of its corresponding
rule instantiation and access the location of an atom occurring in this rule instantiation in constant time.
Algorithm~$\mathsf{A}_{\text{bsc}}$ is then given as follows.\\

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any predicate in $\textbf{I}$, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instantiations in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instantiations is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{\text{bsc}}$ works on the parallel condition,
each processor can check the applicability of its corresponding
rule instantiation can then create edges between atoms in constant time. Thus,
one iteration of \ref{alg1:updateG} costs constant time (denoted by $c_2$). In summary,
the whole computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$ turns out to be $c_1+\ell\cdot c_2$
where $\ell$ denotes the number of iterations of \ref{alg1:updateG}.

@@

We now analyse the parallel complexity of Algorithm~$\mathsf{A}_{\text{opt}}$.
We use the same symbols $e$, $v$ and $r$ that are used for analyzing the complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$
(see Section~\ref{sec:alg-bsc}). Similarly to the analysis for Algorithm~$\mathsf{A}_{\text{bsc}}$, it can be checked that
the number of processors is polynomial in $|\textbf{I}|$, i.e., $r(|\textbf{I}|e)^v$.
We now consider the computing time of Algorithm~$\mathsf{A}_{\text{opt}}$. It
is obvious that \ref{alg3:addFacts} and
\ref{alg3:halt} occupy constant time (denoted by $c_1$). In \ref{alg3:updateG}, the phases of
computing $S_{\!\textit{\tiny rch}}$ and updating $\mathcal{G}$ also take constant time (denoted by $c_2$)
on the parallel condition. Recall that the number of constants is at most $|\textbf{I}|e$ (see Section~\ref{sec:alg-bsc}).
Let $w$ and $p$ represent the maximum arity of any predicate and the number of predicates, respectively.
We have that the number of all possible substitutions of atoms is at most $p(|\textbf{I}|e)^w$.
The relation $S_{\!\textit{\tiny rch}}$ consists of pairs of the atoms. Thus, the size of $S_{\!\textit{\tiny rch}}$
is bounded by $p^2(|\textbf{I}|e)^{2w}$. It can be checked that the computing time of the \texttt{NC} algorithm for computing the transitive
closure of $S_{\!\textit{\tiny rch}}$ is poly-logarithmic in the size of $\textbf{I}$, more precisely,
it is upper bounded by $\log^2(p^2(|\textbf{I}|e)^{2w})$ ($=4\log^2(p)+4w^2\log^2(|\textbf{I}|e)+8w\log(p)\log(|\textbf{I}|e)$).
We now have that the total computing time of
Algorithm~$\mathsf{A}_{\text{opt}}$ is $c_1+\ell c_3 + \ell
c_4\log(|\textbf{I}|e) +\ell c_5\log^2(|\textbf{I}|e)$
where $c_3=c_2+4\log^2(p)$, $c_4=8w\log(p)$, $c_5=4w^2$ and $\ell$ denotes the number of iterations of \ref{alg3:updateG}.
